---
title: "Reproducible Research - Project 1"
author: "Brad Dietz"
date: "November 11, 2015"
output: html_document
---

This project will teach markdown, GitHub and many facets of general R usage using data from a personal activity monitoring device.  The device collected the number of steps over 2 months at 5 minute increments. 

###Loading and preprocessing the data
```{r, cache=TRUE, echo=TRUE}
activity <-read.csv("activity.csv", header = TRUE, sep=",", check.names=FALSE, stringsAsFactors=FALSE)
activity$steps <- as.numeric(activity$steps)
```
Please note that the assignment says to ignore the missing values so rows with NA values will not be dropped.

###What is the mean total number of steps taken per day?

####Total Number of Steps per day
```{r, cache=TRUE, echo=TRUE}
sum(activity$steps, na.rm=TRUE)
```

####Histogram of total steps per day
```{r, echo=TRUE}
library(plyr)
DailySteps<-ddply(activity, c("date"), summarise, Sum=sum(steps, na.rm = T))
hist(DailySteps$Sum,  breaks = 10, main="Histogram of total steps per day", 
     xlab="Total steps per day")
```

####Mean total Number of Steps per day
```{r, cache=TRUE, echo=TRUE}
mean(DailySteps$Sum)
```

####Median total Number of Steps per day
```{r, cache=TRUE, echo=TRUE}
median(DailySteps$Sum)
```

###What is the average daily activity pattern?

####Time series plot 
```{r, echo=TRUE}
library(ggplot2)
library(plyr)
DailyActivity<-ddply(activity, c("interval"), summarise, Mean=mean(steps, na.rm = T))
plot(DailyActivity$interval, DailyActivity$Mean, type = "l",  col=1, 
     main="Average steps averaged across all days", xlab="Interval", 
     ylab="Average steps")
```

####Interval with the maximum number of steps
```{r, echo=TRUE}
DailyActivity[which.max(DailyActivity$Mean),]$interval
```

###Imputing missing values

####Total number of missing values in the dataset
```{r, cache=TRUE, echo=TRUE}
count(is.na(activity$steps))[2,]
```

####Impute values for the NA values in the dataset
Use the MICE package to replace the NA values
```{r, cache=TRUE, echo=TRUE}
library(mice)

ImputedActivity <- mice(activity)
ImputedActivity<-complete(ImputedActivity)
```

####Histogram of total steps per day using the original and imputed datasets
```{r, cache=TRUE, echo=TRUE}
library(plyr)
ImputedDailySteps<-ddply(ImputedActivity, c("date"), summarise, Sum=sum(steps, na.rm = T))

par(mfrow = c(1, 2))
hist(DailySteps$Sum,  breaks = 10, main="Original Histogram", 
     xlab="Total steps per day")
hist(ImputedDailySteps$Sum,  breaks = 10, main="Imputed Histogram", 
     xlab="Total steps per day")
```

####Mean total Number of Steps per day w/imputed dataset
```{r, cache=TRUE, echo=TRUE}
mean(ImputedDailySteps$Sum)
```
The imputed dataset mean is higher than the original dataset mean by 
```{r, cache=TRUE, echo=TRUE}
mean(ImputedDailySteps$Sum)-mean(DailySteps$Sum)
```

####Median total Number of Steps per day w/imputed dataset
```{r, cache=TRUE, echo=TRUE}
median(ImputedDailySteps$Sum)
```
The imputed dataset median is higher than the original dataset median by 
```{r, cache=TRUE, echo=TRUE}
median(ImputedDailySteps$Sum)-median(DailySteps$Sum)
```

####Total Number of Steps per day w/imputed dataset
```{r, cache=TRUE, echo=TRUE}
sum(ImputedDailySteps$Sum)
```
The imputed dataset step total is higher than the original dataset step total by 
```{r, cache=TRUE, echo=TRUE}
sum(ImputedDailySteps$Sum)-sum(DailySteps$Sum)
```

###Are there differences in activity patterns between weekdays and weekends?

####Create a 2 level factor variable (weekday/weekend) in the imputed dataset
```{r, cache=TRUE, echo=TRUE}
library(ggplot2)
ImputedActivity$date <- as.Date(ImputedActivity$date)

myday <- function(x){
  if(x=="Monday")
   return("weekday")
  if(x=="Tuesday")
    return("weekday")
  if(x=="Wednesday")
    return("weekday")
  if(x=="Thursday")
    return("weekday")
  if(x=="Friday")
    return("weekday")
  if(x=="Saturday")
    return("weekend")
  if(x=="Sunday")
    return("weekend")
  else
    return(NA)
}

ImputedActivity$fday <- sapply(weekdays(ImputedActivity$date), myday)
ImputedActivity$fday <- as.factor(ImputedActivity$fday)
```

####Make a time series panel plot of Weekday and Weekend step activity 
```{r, cache=TRUE, echo=TRUE}
WeekActivity<-ddply(ImputedActivity, c("interval", "fday"), summarise, Mean=mean(steps, na.rm = T))

par(mfrow = c(1, 1))
ggplot(WeekActivity, aes(x=interval, y=Mean)) +
geom_line() +    
facet_grid(fday ~ .) +
ylab("Average steps") +
ggtitle("Average steps averaged across weekday/weekend")
```



